---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ibdplexus 

<!-- badges: start -->
<!-- badges: end -->

## Overview

ibdplexus is designed to work with data downloaded from the [IBD Plexus Portal](https://ibdplexus.org) from the Crohn's & Colitis Foundation. It includes functions for the cohorts RISK, SPARC and QORUS. More information on Plexus and the cohorts can be found [here](https://www.crohnscolitisfoundation.org/research/current-research-initiatives/ibd-plexus). Functions include:  

* `load_data()` loads unzipped data from Plexus
* `load_zipped_data()` loads zipped data from Plexus
* `risk_extract_clinical_data()` creates a summary table by visit for RISK 
* `sparc_summary()` creates a summary table for SPARC 
* `calculate_disease_scores()` calculates disease scores for SPARC 
* `sparc_scores()` calculates disease scores at a specific time for SPARC
* `sparc_medication()` predicts the medication a SPARC participant is on at a specific time
* `emr_extract_diagnosis()` subsets the plexus extract based on specified ICD10 codes

## Installation

You can install the released version of ibdplexus from [github](https://github.com/ccf-tfehlmann/ibdplexus) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("ccf-tfehlmann/ibdplexus")
```


## User Guides

* <a href="https://github.com/ccf-tfehlmann/ibdplexus/blob/master/userguides/Calculating_Disease_Activity_Scores_for_SPARC.pdf">`calculate_disease_scores()`</a>  

* <a href="https://github.com/ccf-tfehlmann/ibdplexus/blob/master/userguides/SPARC_Summary_userguide_updated_102221.pdf">`sparc_summary()`</a>  


* <a href="https://github.com/ccf-tfehlmann/ibdplexus/blob/master/userguides/sparc_scores_at_index_userguide.pdf">`sparc_scores()`</a>  

* <a href="https://github.com/ccf-tfehlmann/ibdplexus/blob/master/userguides/RISK_Extract_Clinical_Data_User_Guide_v1.1.pdf">`risk_extract_clinical_data()`</a>  

* <a href="https://github.com/ccf-tfehlmann/ibdplexus/blob/79168cbffa523ab8da142d250133e70f625e00c7/userguides/Diagnosis_EMR_Extract.md">`Diagnosis_EMR_Extract`</a>  


## Example

```{r example, eval = FALSE}
library(ibdplexus)
# 
# loading unzipped SPARC Encounter, Demographics and Prescriptions CRF data only ----
# data = load_data(datadir = ".",cohort = "SPARC", domains = c("Demographics", "Prescriptions", "Encounter"), data_type = "CRF")


## sparc sumamary at the time of endoscopy (+/- 30 days) with indication ----
#
#   data = load_data(datadir = "~/r_input/",cohort = "SPARC", domains = c("ALL"), data_type = "BOTH")
#
#
#   summary_at_endo = sparc_summary(data = data,
#                              index_info = "endoscopy",
#                              filename = "SPARC_SUMMARY_AT_ENDOSCOPY.xlsx")

```
