---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ibdplexus 

<!-- badges: start -->
<!-- badges: end -->

## Overview

ibdplexus is designed to work with data downloaded from the [IBD Plexus Portal](https://ibdplexus.org) from the Crohn's & Colitis Foundation. More information on Plexus and the cohorts can be found [here](https://www.crohnscolitisfoundation.org/research/current-research-initiatives/ibd-plexus). Functions include:  

* `load_data()` loads unzipped data from Plexus
* `load_zipped_data()` loads zipped data from Plexus
* `risk_extract_clinical_data()` creates a summary table by visit for RISK 
* `sparc_summary()` creates a summary table for SPARC 
* `calculate_disease_scores()` calculates disease scores for SPARC 
* `sparc_scores()` calculates disease scores at a specific time for SPARC
* `sparc_medication()` predicts the medication a SPARC participant is on at a specific time
* `emr_extract_diagnosis()` subsets the plexus extract based on specified ICD10 codes

## Installation

You can install the released version of ibdplexus from [github](https://github.com/ccf-tfehlmann/ibdplexus) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("ccf-tfehlmann/ibdplexus")
```


## User Guides

* <a href="https://github.com/ccf-tfehlmann/ibdplexus/blob/master/inst/userguides/Calculating_Disease_Activity_Scores_for_SPARC.pdf">`calculate_disease_scores()`</a>  

* <a href="https://github.com/ccf-tfehlmann/ibdplexus/blob/master/inst/userguides/SPARC_Summary_userguide_updated_102221.pdf">`sparc_summary()`</a>  


* <a href="https://github.com/ccf-tfehlmann/ibdplexus/blob/master/inst/userguides/sparc_scores_at_index_userguide.pdf">`sparc_scores()`</a>  

* <a href="https://github.com/ccf-tfehlmann/ibdplexus/blob/master/inst/userguides/RISK_Extract_Clinical_Data_User_Guide_v1.1.pdf">`risk_extract_clinical_data()`</a>  

* <a href="https://github.com/ccf-tfehlmann/ibdplexus/blob/79168cbffa523ab8da142d250133e70f625e00c7/inst/userguides/Diagnosis_EMR_Extract.md">`Diagnosis_EMR_Extract`</a>  


## Example

```{r example, eval = FALSE}
library(ibdplexus)

# Load in all data in IBD Plexus for the SPARC cohort ----
data <- load_data(datadir = "~/r_input/", cohort = "SPARC", domains = "ALL", data_type = "BOTH")


# Get Excel Tables and Data.Frames with Data Summarised within 60 days of the Enrollment Time Point ----

# SPARC Summary Table
e_sum <- sparc_summary(
  data = data,
  index_info = "ENROLLMENT",
  filename = "SPARC_SUMMARY_ENROLLMENT.xlsx",
  index_range = "60"
)


# SPARC Scores

e_scores <- sparc_scores(
  data = data,
  index_info = "ENROLLMENT",
  filename = "SPARC_SCORES_ENROLLMENT.xlsx",
  index_range = "60"
)


# SPARC Medication

e_med <- sparc_medication(
  data = data,
  index_info = "ENROLLMENT",
  filename = "SPARC_MEDICATION_ENROLLMENT.xlsx"
)
```
