---
title: "Electronic Medical Data in SPARC IBD and IBD QOrus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{emr-data-in-sparc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

Two studies in IBD Plexus have integrated electronic medical record (EMR) data:

1.  A Study of a Prospective Adult Research Cohort with Inflammatory Bowel Disease (SPARC IBD) is a longitudinal study following adult IBD patients as they receive care at 17 different sites across the United States. To learn more about SPARC IBD and it's development please see [The Development and Initial Findings of A Study of a Prospective Adult Research Cohort with Inflammatory Bowel Disease (SPARC IBD)] ([https://doi.org/10.1093/ibd/izab071).](https://doi.org/10.1093/ibd/izab071).)

2.  IBD Qorus is a prospective cohort focused on improving the quality of care for patients with IBD. Currently there are over 30 sites enrolling patients across the United States. To learn more about IBD Qorus please see <https://www.crohnscolitisfoundation.org/research/ibd-qorus>.

This vignette details the functions built into the ibdplexus package to help navigate the EMR data for these two studies.

```{r setup}
library(ibdplexus, quietly = T)
library(tidyr, quietly = T)
library(dplyr, quietly = T)
library(lubridate, quietly = T)
```

# Calculating BMI

`calculate_bmi()` filters the observations_emr domain to find the weight and height of a patient. Weight is converted to kg and height is converted to meters. For patients with more than one BMI, outliers are removed. This function produces a long data.frame with one row per patient BMI.

In the `sparc_summary()` function, this data.frame is cut to find the BMI closest to the specified index date.

Future versions of this package will include a similar functionality for a Qorus summary table.

# Extracting Diagnosis Codes from the EMR

`emr_extract_diagnosis()` filters the diagnosis and patient_problem table based on specified inclusion and exclusion criteria.

This function can search through the EMR using ICD10 codes or keywords. The default function will search through the EMR using ICD10 codes. One may search for comorbidities, general disease complications, common symptoms, extra-intestinal manifestations, cancer, common clinical trial criteria, and infectious diseases using the pre-set condition categories outlined below. There is also an option to custom input ICD10 codes.

The text search option of this function filters through EMR diagnosis descriptions for any keyword. All relevant diagnosis will be returned for the keyword inputted. Diagnosis descriptions that include the specified keyword but are not a positive diagnosis for the patient will not be included. For example, rows are dropped if they are a negative diagnosis ("without" diagnosis), are screenings or scans for a diagnosis, are family history of the diagnosis, or are indicating risk for the diagnosis.

There is also the option to use both codes and text to search through the EMR for a diagnosis. More specifics on the inputs for code and text search options is detailed in the emr_extract_diagnosis parameters section of this document.

The result is a list of data.frames with all relevant EMR ICD10 diagnosis, patient problem and/or patient problem records.

The list of pre-programmed ICD10 and ICD9 codes is below. The text search term column specifies the words and phrases for the pre-programmed categories when both text_search and code_search are true. The text search option will search for all of the full phrase or words in quotations.

| Condition Categories                     | ICD 10 Diagnosis Codes                                                                                                                                                                                         | ICD 9 Diagnosis Codes                                                                                                       | Text Search Terms                                                                                                                                                                                                                                                                                                                                                                                                                                           |
|--------------|-----------------------------|----------------|--------------|
| SYSTEMIC FUNGAL INFECTION                | B39, B45, B38, B40, B46, B44, B37, B59                                                                                                                                                                         | 110-118                                                                                                                     | "SYSTEMATIC FUNGAL INFECTION"                                                                                                                                                                                                                                                                                                                                                                                                                               |
| CANCER                                   | All C codes, d0, d1, d2, d3, d4                                                                                                                                                                                | 140-229                                                                                                                     | "CANCER", "MALIGNANT NEOPLASM", "NEOPLASM OF UNCERTAIN BEHAVIOR"                                                                                                                                                                                                                                                                                                                                                                                            |
| COLORECTAL CANCER                        | C18, C19, C20                                                                                                                                                                                                  | 153, 154.0, 154.1                                                                                                           | "COLORECTAL CANCER", "COLON ADENOCARCINOMA", "MALIGNANT NEOPLASM OF COLON", "RECTAL CANCER", "MALIGNANT NEOPLASM OF SIGMOID COLON", "MALIGNANT NEOPLASM OF RECTOSIGMOID", "RECTOSIGMOID CANCER", "ADENOCARCINOMA OF COLON", "CANCER OF COLON", "CANCER OF SIGMOID COLON", "MALIGNANT NEOPLASM OF DESCENDING COLON" , "COLON CANCER", "MALIGNANT NEOPLASM OF RECTUM", "MALIGNANT TUMOR OF RECTUM", "RECTAL CANCER", "MALIGNANT NEOPLASM OF TRANSVERSE COLON" |
| CERVIX CARCINOMA                         | D06                                                                                                                                                                                                            | 233.1                                                                                                                       | "CERVIX CARCINOMA"                                                                                                                                                                                                                                                                                                                                                                                                                                          |
| SKIN CARCINOMA                           | C44.01, C44.02, C44.11, C44.12, C44.21, C44.22, C44.31, C44.32, C44.41, C44.42, C44.51, C44.52, C44.61, C44.62, C44.71, C44.72, C44.81, C44.82, C44.91, C44.92                                                 | 232                                                                                                                         | "SKIN CARCINOMA"                                                                                                                                                                                                                                                                                                                                                                                                                                            |
| STOMA                                    | L24.B0, L24.B1, L24.B3, Z93.3, Z93.2                                                                                                                                                                           | V44.2, V44.3                                                                                                                | "STOMA", "ILEOSTOMY", "COLOSTOMY"                                                                                                                                                                                                                                                                                                                                                                                                                           |
| DEMYELINATING DISORDER                   | G35, G36, G37                                                                                                                                                                                                  | 340, 341                                                                                                                    | "DEMYELINATING DISORDER", "MULTIPLE SCLEROSIS", "DEMYELINATING DISEASE"                                                                                                                                                                                                                                                                                                                                                                                     |
| CELIAC                                   | K90.0                                                                                                                                                                                                          | 579.0                                                                                                                       | "CELIAC"                                                                                                                                                                                                                                                                                                                                                                                                                                                    |
| PSC                                      | K83.01                                                                                                                                                                                                         | 576.1                                                                                                                       | "PSC", "PRIMARY SCLEROSING CHOLANGITIS"                                                                                                                                                                                                                                                                                                                                                                                                                     |
| GI BLEEDING                              | K92.1                                                                                                                                                                                                          | 569.3, 578.1, 599.70, 777.3, 792.1                                                                                          | "GI BLEEDING", "BLOOD IN STOOL", "HEMATOCHEZIA", "MELENA", "BLOODY STOOLS", "GASTROINTESTINAL HEMORRHAGE", "BLACK TARRY STOOLS"                                                                                                                                                                                                                                                                                                                             |
| GI ULCER                                 | K25, K27, K28, K26, K63.3, K62.6                                                                                                                                                                               | 531-534                                                                                                                     | "GI ULCER", "ULCER OF ILEUM", ULCER OF ANUS", "UCLER OF RECTUM"                                                                                                                                                                                                                                                                                                                                                                                             |
| PERIANAL ABSCESS OR FISTULA              | K50.913, K50.914, K50.813, K50.814, K50.013, K50.014, K50.113, K50.114, K51.013, K51.014, K51.213, K51.214, K51.313, K51.314, K51.413, K51.414, K51.513, K51.514, K51.813, K51.814, K51.913, K51.914, K60, K61 | 565.1, 566                                                                                                                  | "PERIANAL ABSCESS OR FISTULA"\*                                                                                                                                                                                                                                                                                                                                                                                                                             |
| WEIGHT LOSS                              | R63.4                                                                                                                                                                                                          | 783.1, 783.2                                                                                                                | "WEIGHT LOSS", "LOSS OF WEIGHT"                                                                                                                                                                                                                                                                                                                                                                                                                             |
| B2 OR B3                                 | K50.912, K50.112, K50.012, K50.812                                                                                                                                                                             | 560.89, 560.9                                                                                                               | "CROHN'S DISEASE WITH INTESTINAL OBSTRUCTION", "CROHN'S DISEASE OF SMALL INTESTINE WITH INTESTINAL OBSTRUCTION", "CROHN DISEASE, WITH INTESTINAL OBSTRUCTION", "CROHN'S DISEASE OF BOTH SMALL AND LARGE INTESTINE WITH INTESTINAL OBSTRUCTION", "CROHN'S DISEASE OF ILEUM WITH INTESTINAL OBSTRUCTION", "CROHN'S DISEASE OF LARGE INTESTINE WITH INTESTINAL OBSTRUCTION"                                                                                    |
| MALNOURISHMENT                           | E4                                                                                                                                                                                                             | 263.9, 269.9                                                                                                                | "MALNOURISHMENT", "ALIMENTARY EDEMA", "MALNUTRITION"                                                                                                                                                                                                                                                                                                                                                                                                        |
| ANEMIA                                   | D50, D51, D52, D53                                                                                                                                                                                             | 280-281                                                                                                                     | "ANEMIA"                                                                                                                                                                                                                                                                                                                                                                                                                                                    |
| DIARRHEA                                 | R19.7, K59.1, K58.0                                                                                                                                                                                            | 564.5, 787.91                                                                                                               | "DIARRHEA"                                                                                                                                                                                                                                                                                                                                                                                                                                                  |
| NAUSEA OR VOMITING                       | R11                                                                                                                                                                                                            | 787.0                                                                                                                       | "NAUSEA", "VOMITING"                                                                                                                                                                                                                                                                                                                                                                                                                                        |
| HYPOALBUMINEMIA                          | E88.09                                                                                                                                                                                                         | 273.8                                                                                                                       | "HYPOALBUMINEMIA"                                                                                                                                                                                                                                                                                                                                                                                                                                           |
| FEVER                                    | R50.9, R61                                                                                                                                                                                                     | 780.6                                                                                                                       | "FEVER", "FEBRILE"                                                                                                                                                                                                                                                                                                                                                                                                                                          |
| ABDOMINAL PAIN                           | R10                                                                                                                                                                                                            | 789.0                                                                                                                       | "ABDOMINAL PAIN", "DYSPEPSIA", "FLANK PAIN", "PELVIC PAIN", "EPIGASTRIC PAIN", "ABDOMINAL DISCOMFORT", "ABDOMINAL CRAMPING", "ACUTE ABDOMEN"                                                                                                                                                                                                                                                                                                                |
| CDI                                      | A04.7                                                                                                                                                                                                          | 008.45                                                                                                                      | "CLOSTRIDIUM DIFFICILE", "C. DIFFICILE"                                                                                                                                                                                                                                                                                                                                                                                                                     |
| ARTHRITIS OR LOW BACK PAIN               | M13, M05, M06, M07, M08, M10, M11, M12, M14, M1A, ( need to add M54.5)                                                                                                                                         | 710-716, 724.2                                                                                                              | "ARTHRITIS", "LOW BACK PAIN", "ARTHOPATHY"                                                                                                                                                                                                                                                                                                                                                                                                                  |
| DACTYLITIS                               | L08.9                                                                                                                                                                                                          | 686.9                                                                                                                       | *TBD*                                                                                                                                                                                                                                                                                                                                                                                                                                                       |
| NON UC IBD DIAGNOSIS                     | K50, K52.3, K52.83, K55.9                                                                                                                                                                                      | 555, 558.9                                                                                                                  | "NON UC IBD DIAGNOSIS", "CROHN'S", "IBD UNCLASSIFIED", "INDETERMINATE COLITIS"                                                                                                                                                                                                                                                                                                                                                                              |
| TOXIC MEGACOLON                          | K59.31                                                                                                                                                                                                         | 564.7                                                                                                                       | "TOXIC MEGACOLON"                                                                                                                                                                                                                                                                                                                                                                                                                                           |
| FULMINANT COLITIS                        | K55.03                                                                                                                                                                                                         | 557.0                                                                                                                       | "FULMINANT COLITIS"                                                                                                                                                                                                                                                                                                                                                                                                                                         |
| INTRAABDOMINAL ABSCESS                   | L02.211, K65.1                                                                                                                                                                                                 | 567.22, 682.2                                                                                                               | "INTRAABDOMINAL ABSCESS", "ABSCESS OF ABDOMINAL", ABDOMINAL WALL ABSCESS", "MESENTERIC ABSCESS", "PERITONEAL ABSCESS", "INTRA-ABDOMINAL ABSCESS"                                                                                                                                                                                                                                                                                                            |
| STRICTURE STENOSIS                       | K56.69                                                                                                                                                                                                         | 560.89                                                                                                                      | "STRICTURE", "STENOSIS"                                                                                                                                                                                                                                                                                                                                                                                                                                     |
| COLON ADENOMA                            | D12.2, D12.3, D12.4, D12.5, D12.6, K31.A2, K55.20                                                                                                                                                              | 211.3, 235.2                                                                                                                | "COLON ADENOMA", "BENIGN NEOPLASM OF COLON", "DYSPLASIA OF COLON", "ADENOMA OF COLON", "COLON DYSPLASIA", "COLONIC ADENOMA"                                                                                                                                                                                                                                                                                                                                 |
| INFECTION                                | L0, A49, A0 (need to refine and add B99)                                                                                                                                                                       | 001-009,130-136                                                                                                             | "INFECTION"                                                                                                                                                                                                                                                                                                                                                                                                                                                 |
| TUBERCULOSIS                             | A15, A17, A18, A19                                                                                                                                                                                             | 010-018                                                                                                                     | "TUBERCULOSIS", "TB INFECTION"                                                                                                                                                                                                                                                                                                                                                                                                                              |
| DIABETES                                 | E08, E09, E10, E11, E13                                                                                                                                                                                        | 250                                                                                                                         | "DIABETES"                                                                                                                                                                                                                                                                                                                                                                                                                                                  |
| HYPERTENSION                             | I10, I11, I12, I13, I15, I16                                                                                                                                                                                   | 401-405                                                                                                                     | "HYPERTENSION", "HTN"                                                                                                                                                                                                                                                                                                                                                                                                                                       |
| COPD                                     | J44                                                                                                                                                                                                            | 491.21, 493.2, 496                                                                                                          | "COPD", "CHRONIC OBSTRUCTIVE PULMONARY DISEASE", "CROHNIC BRONCHITIS", "CHRONIC AIRWAY OBSTRUCTION", "CHRONIC OBSTRUCTIVE ASTHMA"                                                                                                                                                                                                                                                                                                                           |
| CKD STAGE IIB OR MORE                    | N18.32, N18.4, N18.5                                                                                                                                                                                           | 585.2-585.5                                                                                                                 | "CKD", "CHRONIC KIDNEY DISEASE", "CHRONIC RENAL FAILURE"                                                                                                                                                                                                                                                                                                                                                                                                    |
| UNSTABLE ANGINA OR MYOCARDIAL INFARCTION | I20, I21                                                                                                                                                                                                       | 410, 412, 413                                                                                                               | "UNSTABLE ANGINA", "MYOCARDIAL INFARCTION", "ANGINA PECTORIS"                                                                                                                                                                                                                                                                                                                                                                                               |
| AUTOIMMUNE INFLAMMATORY DISEASE          | M05, M06, M3, M04                                                                                                                                                                                              | 714, 710                                                                                                                    | "AUTOIMMUNE INFLAMMATORY DISEASE"                                                                                                                                                                                                                                                                                                                                                                                                                           |
| HEPATITIS B                              | B16, B18.0, B18.1, B19.1                                                                                                                                                                                       | 070.2, 070.3                                                                                                                | "HEPATITIS B"                                                                                                                                                                                                                                                                                                                                                                                                                                               |
| HEPATITIS C                              | B17.1, B18.2, B19.2                                                                                                                                                                                            | 070.41, 070.44, 070.51, 070.54, 070.7                                                                                       | "HEPATITIS C"                                                                                                                                                                                                                                                                                                                                                                                                                                               |
| INHERITED AUTOIMMUNE DISORDER            | D80.0, D82.0, D80.4, D82.3, N41.4, Q82.8, D81.0, D81.1, D81.2, D81.3, E70.330, D76.1, D82.4, D82.2, D81.6, D81.7, D83, D80.2, D84.1, G11.3, D81.5, D81.8                                                       | 279.04, 279.12, 279.02, 279.8, 601.8, 757.2, 757.39, 279.2, 270.2, 288.4, 279.8, 279.06, 279.01, 277.6, 334.8, 277.2, 266.2 | "INHERITED AUTOIMMUNE DISORDER"                                                                                                                                                                                                                                                                                                                                                                                                                             |

### emr_extract_diagnosis parameters

`data` is a list of data frames typically generated by load_data from the ibdplexus package. It must include demographics and the diagnosis, patient problem and/or patient history tables. Only EMR data is required.

`code_search` is the option to extract relevant diagnosis codes using ICD 10 codes. This defaults to true in the function. If code_search is true, the inclusion parameter must be either one of the pre-set condition categories from above, or "CUSTOM". The function will not run if code_search is true and an unknown character string is in the inclusion criteria.

`text_search` is the option to extract relevant diagnosis codes using keyword text search. This defaults to false. When set to true and code_search is false, the inclusion parameter can be any string of characters. The different arguments of the character string should be separated by a pipe (eg. "argument1\|argument2").

`inclusion` When searching by code, the diagnosis category of ICD 10 codes to search for. Can choose between predefined categories outlined above or enter a custom defined list. If searching for a custom defined list, this parameter must be set to "CUSTOM", and the list should be specified in the `custominc` parameter. When text searching, a regular expression string of keywords to search for (eg. "argument1\|argument2\|argument3")

**note for combined text and code search**: The function allows for code_search and text_search to both be true, so the function will filter through the EMR using both ICD codes and keyword text search. Currently, the function will only run with both of these options if the input is one of the pre-programmed condition categories. For example, if text_search = T and code_search = T with inclusion = "COLORECTAL CANCER", the function will return all results that match the ICD 10 & ICD 9 codes in the colorectal cancer condition category, as well as any results that have any of the specified text strings in the description of the diagnosis but are not one of the specified codes (ie. a SNOMED code might return for a combined text and code search).

`exclusion` diagnosis category of ICD 10 codes to exclude from search. Can currently be either "SKIN CARCINOMA", "CERVIX CARCINOMA", or "CUSTOM" (for which the custom exclusion list will be specified in the `customexc` parameter). More predefined lists can be included in future updates of the function.

`custominc` optional list of custom ICD 10 codes to be searched for. The variable "inclusion" must also be set to "CUSTOM" for this option. If multiple inputs, needs to be in a regular expression string (eg. "argument1\|argument2\|argument3")

`customexc` optional list of custom ICD 10 codes to be excluded from the search. The variable "exclusion" must also be set to "CUSTOM" for this option. If multiple inputs, needs to be in a regular expression string (eg. "argument1\|argument2\|argument3")

`datecutoff` Type of date cutoff to be used for EMR extract. Current options include "ENROLLMENT" - number of days before consent into study.

`index_range` Number of days before date cutoff to be used for EMR extract. Time must be specified in number of days. Default is 36500 days (100 years).
