---
title: "Medication Data in SPARC IBD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{medication-in-SPARC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)

options(rmarkdown.html_vignette.check_title = FALSE)


```

A Study of a Prospective Adult Research Cohort with Inflammatory Bowel Disease (SPARC IBD) is a longitudinal study following adult IBD patients as they receive care at 17 different sites across the United States. To learn more about SPARC IBD and it's development please see [The Development and Initial Findings of A Study of a Prospective Adult Research Cohort with Inflammatory Bowel Disease (SPARC IBD)] (<https://doi.org/10.1093/ibd/izab071>).

Data from patient reported surveys (eCRFs), IBD Smartform and electronic medication records (EMR) are integrated into IBD Plexus. The ibdplexus package was created to synthesize this data into research ready formats. This vignette focuses on the functions in the ibdplexus package that handle the medication information in SPARC IBD.

Patient reported surveys, sent out at enrollment and quarterly, ask patients about their current IBD medication use specifically. Electronic medical records, when available, contain the prescriptions for each patient from that institution. They may not be prescribed specifically for IBD. The Smartform, updated at the time of office visit by the provider, asks for exposure to specific IBD medications.

This vignette mainly focuses on the medication data from eCRF and EMR.

# Filtering for Medications of Interest

`sparc_med_filter` filters through the eCRF and EMR data for medications of interest that are listed in the data.frame `med_grp`.

`med grp` contains the medication name of interest, the type of medication and a standardized name. The MEDICATION_NAME column includes generic and brand names. the new_med_name column is the standardized name for each MEDICATION_NAME. Currently, all biosimilars are mapped to the generic name.

A future iteration will parse out biosimilars into their own group.

```{r eval = TRUE, warning = FALSE, message = FALSE}

library(ibdplexus, quietly = T)
library(tidyr, quietly = T)
library(dplyr, quietly = T)
library(lubridate, quietly = T)

# Here is the table of the medication types in med_grp
knitr::kable(med_grp %>% distinct(med_type)) 

# Here is the first 5 rows of the med_grp data.frame

knitr::kable(head(med_grp, 5)) 

# Here is how Infliximab and its' bio-similars are mapped 

knitr::kable(med_grp %>% filter(new_med_name == "Infliximab"))

```

To see the entire med_grp data.frame click <a href="https://github.com/ccf-tfehlmann/ibdplexus/blob/79168cbffa523ab8da142d250133e70f625e00c7/data/med_grp.rda">`here`</a>

`sparc_med_filter` requires the following SPARC domains to be loaded:

-   prescriptions

-   observations

-   demographics

-   encounter

One must also supply the med_groups of interest which corresponds to the med_type in the med_grp data.frame.

`sparc med filter` subsets the prescriptions table to just the medication types specified in med_groups. The leading question on medications is located in the observations table.

In the baseline survey, a patient is asked if they are currently taking any medications for their IBD. If they answer "No" they are flagged as having no medications at enrollment. It is important to note that antibiotics and probiotics are asked about separately in the patient surveys.

In the quarterly follow-up surveys, a patient is asked if they have had any changes in their IBD medication. If a patient indicates they have not had any medication changes, the data from the previous survey is pulled forward.

`sparc med filter` uses grep to search through the MEDICATION, OTHER_MEDICATION and SRC_DRUG_CODE_CONCEPT_NAME columns in the prescription table. If the medication of interest is found, it subsets the data and maps the medication to the standardized name (new_med_name).

```{r}

# Load EMR and eCRF data necessary for sparc_med_filter 

data <- load_data(datadir = "~/r_input/", cohort = "SPARC", domains = c("prescriptions", "observations", "demographics", "encounter"), data_type = "Both")

# Filter SPARC IBD data for biologics & immunomodulators 

meds_of_interest <- sparc_med_filter(
  data$prescriptions,
  data$observations,
  data$demographics,
  data$encounter,
  med_groups = c("Biologic","Immunomodulators")
)

```

# Finding the First Instance of a Medication

`sparc_med_starts` looks at a patients treatment journey utilizing EMR and eCRF data. It is designed to find the first instance of a medication. It is only appropriate for long term medications that we do not expect to re-start. We assume biologics, aminosalicylates and immunomodulators fit this description. It returns a data.frame with the first medication start date for each drug, and an end date if available.

It calls `sparc_med_filter` to filter for the medications of interest. It is recommended to only use those medications that a patient is on long term and unlikely to restart.

Medication start and stop dates are chosen independently from both eCRF and EMR sources. Medications with a start or stop date before 1980 are dropped.

For EMR data, if a medication start date is missing, the visit encounter start date is used. These records are flagged in the column VISIT_ENCOUNTER_MED_START.

If a patient has medication information for the same drug from eCRF and EMR, the eCRF data is preferred and used to generate MED_START_DATE and MED_END_DATE. If only EMR data is available for that medication, then EMR data is used.

Any overlap between medications is reported along with the number of days the medications overlap.

If no end date is given for a prescription, the duration of the overlap is calculated assuming an ongoing prescription. The effective end date is set using a database wide cutoff based on the the date of the latest encounter any patient had (as returned by the function `extract_latest`.

`sparc_med_starts` creates a table with the following columns:

+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Column                                  | Description                                                                                                                                                                                                                                                 |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| DEIDENTIFIED_MASTER_PATIENT_ID          | Unique patient id                                                                                                                                                                                                                                           |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| MEDICATION                              | Corresponds to new_med_name in the med_grp data.frame                                                                                                                                                                                                       |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| MED_START_DATE_ECRF                     | Medication Start Date from the eCRF data source                                                                                                                                                                                                             |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| MED_END_DATE_ECRF                       | Medication End Date from the eCRF data source                                                                                                                                                                                                               |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| CURRENT_MEDICATION_ECRF                 | YES if the medication is current, NO if the medication has been stopped                                                                                                                                                                                     |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| NUMBER_OF_DOSE_CHANGES_ECRF             | Counts the number of times a patient reported a change in dose. This field is free text and the r script only parses out numbers.                                                                                                                           |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| NUMBER_OF_FREQ_CHANGES_ECRF             | Counts the number of times a patient reported a change in medication frequency.                                                                                                                                                                             |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| REASON_STOPPED_ECRF                     | Lists the reason stopped if available. A medication may be current, but have a reason stopped populated because of a dose and/or frequency change.                                                                                                          |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| VISIT_ENCOUNTER_MED_START               | For EMR data, flags if a medication start date is missing and the visit encounter start date was used instead                                                                                                                                               |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| MED_START_DATE_EMR                      | Medication Start Date from the EMR data source                                                                                                                                                                                                              |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| MED_END_DATE_EMR                        | Medication End Date from the EMR data source                                                                                                                                                                                                                |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| LOADING_DOSE_EMR                        | 1 if a patients prescription pattern for specific biologics represents the loading pattern of that medication.                                                                                                                                              |
|                                         |                                                                                                                                                                                                                                                             |
|                                         | +--------------------+---------------------------------------------------------------------------------+                                                                                                                                                    |
|                                         | | Biologic           | Criteria                                                                        |                                                                                                                                                    |
|                                         | +====================+=================================================================================+                                                                                                                                                    |
|                                         | | Infliximab         | Prescriptions at 0,2,6 and then every 8 weeks                                   |                                                                                                                                                    |
|                                         | +--------------------+---------------------------------------------------------------------------------+                                                                                                                                                    |
|                                         | | Adalimumab         | Every 2 weeks with the first dose as 160mg, then 80 mg, then 40mg going forward |                                                                                                                                                    |
|                                         | +--------------------+---------------------------------------------------------------------------------+                                                                                                                                                    |
|                                         | | Certolizumab Pegol | Prescriptions at 0,2,4 and then every 4 weeks                                   |                                                                                                                                                    |
|                                         | +--------------------+---------------------------------------------------------------------------------+                                                                                                                                                    |
|                                         | | VedolizInumab      | Prescriptions at 0,2,6 and then every 8 weeks                                   |                                                                                                                                                    |
|                                         | +--------------------+---------------------------------------------------------------------------------+                                                                                                                                                    |
|                                         | | Ustekinumab        | Initial dose as an IV and then subcutaneous injection every 8 weeks             |                                                                                                                                                    |
|                                         | +--------------------+---------------------------------------------------------------------------------+                                                                                                                                                    |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| PRESCRIPTION_NUMBER_EMR                 | Number of rows for this medication in the prescription emr data                                                                                                                                                                                             |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| MEDICATION_REFILLS_EMR                  | Number of refills from most recent prescription in the EMR data                                                                                                                                                                                             |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ECRF_PRESCRIPTION_DATA                  | 1 if data for this medication is available in the eCRF data source                                                                                                                                                                                          |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| MED_START_DATE                          | If ECRF_PRESCRIPTION_DATA = 1, then from MED_START_DATE_ECRF, else from MED_START_DATE_EMR.                                                                                                                                                                 |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| MED_END_DATE                            | If ECRF_PRESCRIPTION_DATA = 1, then from MED_END_DATE_ECRF, else from MED_END_DATE_EMR.                                                                                                                                                                     |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| MOA                                     | The mechanism of action for each medication. Includes, antiTNF, Aminosalicylates, Immunomodulators, Interleukin-12 and -23 Antagonist, JAKi, Integrin Receptor Antagonists, S1P                                                                             |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| MEDICATION_NUMBER                       | counts the number of different medications in a patients' journey. Only medications in the selected med_groups are considered. The medication with with earliest start date will have MEDICATION_NUMBER = 1.                                                |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| OVERLAPPING_MOA                         | If multiple medications overlap, the overlapping MOA is listed here                                                                                                                                                                                         |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| OVERLAPPING_DAYS                        | The number of days the medication overlaps with another MOA                                                                                                                                                                                                 |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| DATE_OF_CONSENT                         | The date of enrollment into SPARC for the patient                                                                                                                                                                                                           |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| DATE_OF_CONSENT_WITHDRAWN               | The date of withdrawal from SPARC, if applicable                                                                                                                                                                                                            |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| STARTED_AFTER_ENROLLMENT                | is 1 if the medication start date is after the date of consent.                                                                                                                                                                                             |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| NO_CURRENT_IBD_MEDICATION_AT_ENROLLMENT | 1 if a patient indicates "No" to "Are you currently on any IBD medications?" at their baseline survey.                                                                                                                                                      |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ADVANCED_MED                            | is 1 if the record is for an advanced medication including Adalimumab, Certolizumab Pegol, Golimumab, Infliximab, Natalizumab, Other Biologic, Ustekinumab, Vedolizumab, Tofacitinib, Upadacitinib, Ozanimod, or Risankizumab. Biosimilars are included.    |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| FIRST_ADVANCED_MED                      | is 1 if a medication record is the first advanced medication a patient receives.                                                                                                                                                                            |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| FIRST_ADVANCED_MED_NUMBER               |  If a record is the first advanced medication a patient receives, FIRST_BIOLOGIC_NUMBER is equal to the MEDICATION_NUMBER.                                                                                                                                  |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| BIONAIVE                                | is 1 if patient has no prior reported advanced medication use                                                                                                                                                                                               |
+-----------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

```{r}


# Find Medication Start Dates for Biologics and Immunomodulators. 
# If export = TRUE then an excel spreadsheet is generated. 

medication_starts <- sparc_med_starts(
  data$prescriptions,
  data$observations,
  data$demographics,
  data$encounter,
  med_groups = c("Biologic","Immunomodulators"),
  export = TRUE
)


```

# Medication at a Specific Time-point

This medication journey can then be used to find the medications a patient is on at a specific time point using the `sparc_medication` function. This time point is referred to as an "index" date. The function looks to see if the index date is within the medication start and stop dates as generated by the `sparc_med_starts` function.

`sparc_medication` includes pre-programmed index dates for enrollment, latest, endoscopy, omics and biosample collection. A data frame with the DEIDENTIFIED_MASTER_PATIENT_ID and index_date can also be passed into the function. If "Enrollment" is specified as the index, then only the patient reported data (MED_START_DATE_ECRF & MED_END_DATE_ECRF) is used to determine what medication a patient is on as this data is very robust at this time point. If two medications of the same MOA overlap, the start of the 2nd medication is used as the end date of the previous medication.

`sparc_medication` also requires a list of data.frames as generated by `load_data` which must include demographics, diagnosis, encounter, procedures, observations, biosample, omics_patient_mapping, and prescriptions.

Currently, this function does not support steroid or antibiotic use. Functions to parse out this information are in development.

A data.frame and excel file is generated with the following columns:

+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| **Column**                              | **Description**                                                                                                                   |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| DEIDENTIFIED_MASTER_PATIENT_ID          | Unique patient id                                                                                                                 |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| DATE_OF_CONSENT                         | The date of enrollment into SPARC for the patient                                                                                 |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| DATE_OF_CONSENT_WITHDRAWN               | The date of withdrawal from SPARC, if applicable                                                                                  |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| BIRTH_YEAR                              | Year of patients birth                                                                                                            |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| SEX                                     | Male, Female or Unknown                                                                                                           |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| DIAGNOSIS                               | Crohn's Disease, Ulcerative Colitis, or IBD Unclassified                                                                          |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| DIAGNOSIS_DATE                          | Date of IBD Diagnosis                                                                                                             |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| INDEX_DATE                              | Date of interest as specified in "index_info"                                                                                     |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| NO_CURRENT_IBD_MEDICATION_AT_ENROLLMENT | 1 if a patient indicates "No" to "Are you currently on any IBD medications?" at their baseline survey.                            |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| MEDICATION_AT_INDEX                     | A concatenated list of medications the patient is on based on if the index date falls between the medication start and stop date. |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| BIONAIVE                                | is 1 if patient has no prior reported biologic use                                                                                |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| MED_START_DATE_ECRF_XXX                 | The start date in the ECRF for medication XXX                                                                                     |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| MED_END_DATE_ECRF_XXX                   | The end date in the ECRF for medication XXX                                                                                       |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| MED_START_DATE_EMR_XXX                  | The start date in the EMR for medication XXX                                                                                      |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| MED_END_DATE_EMR_XXX                    | The end date in the EMR for medication XXX                                                                                        |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+
| CURRENT_MEDICATION_ECRF_XXX             | The current medication flag (yes/no) from eCRF for medication XXX                                                                 |
+-----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------+

```{r}


# Load Data needed for sparc_medication function 

data <- load_data(datadir = "~/r_input/", cohort = "SPARC", domains = c ("demographics", "diagnosis", "encounter", "procedures", "observations", "biosample", "omics_patient_mapping","prescriptions"), data_type = "Both")

# Find Biologics and Immunomodulators a patient is on at enrollment. 

med_at_enrollment <- sparc_medication(
  data = data,
  index_info = "Enrollment",
  med_groups = c("Biologic", "Immunomodulators"),
  filename = "SPARC_MEDICATION_AT_ENROLLMENT.xlsx"
)

# Find Biologics and Immunomodulators a patient is on at endoscopy

med_at_enrollment <- sparc_medication(
  data = data,
  index_info = "Endoscopy",
  med_groups = c("Biologic", "Immunomodulators"),
  filename = "SPARC_MEDICATION_AT_ENDOSCOPY.xlsx"
)

```
