---
title: "RISK Summary Table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{meds_at_visit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

The [RISK cohort](https://www.crohnscolitisfoundation.org/research/current-research-initiatives/pediatric-risk-stratification) is a closed observational prospective cohort study which enrolled pediatric (\<16 years old) patients within 30 days of a suspected of a Crohn's Disease diagnosis. The final cohort enrolled 1,800 patients across 28 clinics in the U.S. and Canada, with a particular focus on the 913 children with a Crohn's Disease diagnosis who were complication free in the first 90 days post diagnosis. Follow-ups including clinical, demographic and biosample data were conducted on a schedule every 6 months for a total of an 8 year follow-up period.

There are several functions that are used in creating the final RISK summary table. All data used in any RISK summary functions should be loaded using the `ibdplexus` R package `load_data` function.

# RISK Summary Table

```{r eval = TRUE, warning = FALSE, message = FALSE, echo = F}
library(ibdplexus, quietly = T)
library(tidyverse, quietly = T)
library(knitr, quietly = T)
# library(kableExtra, quietly = T)
library(readxl, quietly = T)
```

`risk_summary` creates a table with the following columns:

```{r echo = F}
knitr::kable(simple_risk_columns)
# ibdplexus::simple_risk_columns %>%
#   kbl(booktabs = T) %>%
#   kable_styling(full_width = T, latex_options = c("striped", "scale_down"))
```
